Dotted Suffix Trees\\A Structure for Approximate Text Indexing

### 2005-09-19 paper/conclusion.tex
A submission ready version of the paper

Along the same lines, the structure's definition might be extended to structures such as the suffix-DAG presented in~\cite[7.7]{gusfield:algorithms}. An algorithm to efficiently construct it over these suffix-DAGs is another open problem.

>>>>

Along the same lines, the structure's definition might be extended to structures such as the suffix-DAG presented in~\cite[\S\,7.7]{gusfield:algorithms}. An algorithm to efficiently construct it over these suffix-DAGs is another open problem.

### 2005-09-19 paper/construction.tex
A submission ready version of the paper

\subsection{One-Error Dotted Tree}
\subsubsection{The Algorithm}

@@

\begin{enumerate}
\item It does not have the leaf labeled 1 in the original tree.
\item All other leafs are labeled with value which is the original leaf's value minus one.
\end{enumerate}

It is easy to see why this is so. The leaf labeled 1 would have resulted in the pattern \putstring{.s\$} which does not occur in original string (it is one character too large). For any other leaf $\overline{w\$}$ which occurred at position $p$ in the string, we have a new leaf $\overline{.w\$}$ which occurs at position $p-1$ in the string.

@@

\begin{enumerate}
\item The leaf labeled 1 in the original error tree is not included
\item A leaf labeled $p$ is included only if $s_{p-1}$ equals the label of the suffix link
\item Leafs in the copy have a label which is the original label's value minus one
\end{enumerate}

@@

%\input{copy-subtree-algo}
%
The only point to note is line~\ref{algoline:copy-subtree:merge}. Since we filter some leafs, without a merging procedure, it would be possible to have nodes which possess only one child. These need to be removed.

@@

It is very simple to remove these nodes. As discussed above, suffix trees are normally implemented such that each node is actually just a pair of indices into the original string (exemplified in Figure~\ref{fig:mississippi-nodes}). To merge a child with its parent, we just need to replace both by a node whose starting index is the starting index of the original child minus the length of the parent edge. This process is illustrated in Figure~\ref{fig:merge}.

@@

To add dot links at every node, we need to add the dot link at the root and then for every node call the procedure addDotLink. This is implemented as Algorithm~\ref{algo:construct-1}. The \textit{foreach} loop at line~\ref{algoline:construct-1:foreach} can be implemented using Depth First Search on the tree. In fact, the procedure addDotLink only needs to be called for certain nodes (those which do not possess an incoming suffix link). However, there is no readily available way to access only these nodes and the procedure above does little extra work.

\subsubsection{Time cost}

@@

@@

\input{construct-k-algo}
\input{copy-subtree-with-dot-link-algo}

The code for the copy is Algorithm~\ref{algo:copy-subtree-with-dot-link} which differs from Algorithm~\ref{algo:copy-subtree} in that dot link are also followed while copying the tree.

>>>>

\subsection{One-Error Dotted Tree}

@@

\begin{enumerate}
\item It does not have the leaf labeled 1 in the original tree. Keeping this leaf would have resulted in the \putstring{.s\$} which does not occur in original string (it is one character too large).
\item For any other leaf $\overline{w\$}$ which occurred at position $p$ in the string, we have a new leaf $\overline{.w\$}$ which occurs at position $p-1$ in the string. Therefore, we adjust the labels.

\end{enumerate}

@@

\begin{enumerate}
\item The leaf labeled 1 in the original error tree is not included.
\item Leafs in the copy have a label which is the original label's value minus one.
\item A leaf labeled $p$ is included only if $s_{p-1}$ equals the label of the suffix link.
\end{enumerate}

@@

\input{copy-subtree-algo}

The only point to note is line~\ref{algoline:copy-subtree:merge}. Since we filter some leafs, without a merging procedure, it would be possible to have nodes which possess only one child. These are removed by merging a child with its parent.

@@

As discussed above, suffix trees are implemented such that each node is actually just a pair of indices into the original string (examplified in Figure~\ref{fig:mississippi-nodes}). To merge a child with its parent, we just need to replace both by a node whose starting index is the starting index of the original child minus the length of the parent edge. This process is illustrated in Figure~\ref{fig:merge}.

@@

Algorithm~~\ref{algo:construct-1} constructs the 1-error dotted tree. The \textit{foreach} loop at line~\ref{algoline:construct-1:foreach} can be implemented using Depth First Search. In fact, the procedure addDotLink only needs to be called for certain nodes (those which do not possess an incoming suffix link). However, there is no readily available way to access only these nodes and the procedure above does little extra work.

\subsubsection*{Time cost}

@@

\input{copy-subtree-with-dot-link-algo}

@@

\input{construct-k-algo}

The code for the subtree copy is Algorithm~\ref{algo:copy-subtree-with-dot-link} which differs from Algorithm~\ref{algo:copy-subtree} in that dot link are also followed while copying the tree.

### 2005-09-19 paper/experimental.tex
A submission ready version of the paper

\subsection{Indexing Time and Size}

We report the indexing time and size (as a count of the nodes in the structure) as a function of the number of character (by using larger and larger prefixes of the datasets).

\begin{figure}%
\centering
\includegraphics[width=.7\textwidth]{ratio-in=english-01-12.eps}%
\caption{Ratios of k-error and (k+1)-error dotted trees}\label{fig:ratios}%
\end{figure}

To experimentally verify the average case, we show in Figures~\ref{fig:ratios} the ratios between the k-error and the (k+1)-error dotted trees. We can easily see that for all three data sets, the experimental values do resemble a logarithm as predicted.

@@

For searching, the full texts were first indexed and then online searches were performed on top of the structure. To generate input patterns, randomly chosen bits of the text were taken. These patterns were then altered by substituting a character at random. To generate mismatches, one character too many was altered~\footnote{Since the characters used were not present in the original text, there is no risk of generating a string which \emph{does appear} in the text.}. The search algorithm performed an early exit, without reporting of occurrences (ie, it only reported whether the string exists in the text). Therefore, the number of occurrences had no influence on the search time.

\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{search-vary-N.in=all.eps}
\caption{Search time versus pattern size}\label{fig:search-vary-m}
\end{figure}

Figure~\ref{fig:search-vary-m} shows the search time for growing pattern lenghts. It shows the case where the pattern exists and where it does not (an extra error is added). The fact that the search is done using an early-exit strategy explains why non existing patterns take longer than existing patterns. The value presented is the number of character comparisons the algorithm makes.

>>>>

\subsection{Index Size}

\begin{figure*}%
\centering
\includegraphics[width=.7\textwidth]{ratio-in=english-01-12.eps}%
\caption{Ratio of number of nodes in k-error and (k+1)-error dotted trees}\label{fig:ratios}%
\end{figure*}

To experimentally verify the average case prediction, we show in Figures~\ref{fig:ratios} the ratios between the k-error and the (k+1)-error dotted trees, for english text (the other datasets show similiar behaviour). We can easily see that the experimental values do resemble a logarithm as predicted.

@@

For searching, the full texts were first indexed and then online searches were performed on top of the structure. To generate input patterns, randomly chosen snippets of the text were taken. These patterns were then altered by substituting a character at random. To generate mismatches, one character too many was altered~\footnote{Since the characters used were not present in the original text, there is no risk of generating a string which \emph{does appear} in the text.}. The search algorithm performed an early exit, without reporting of occurrences (ie, it only reported whether the string exists in the text). Therefore, the number of occurrences had no influence on the search time.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{search-vary-N.in=all.eps}
\caption{Search time versus text size}\label{fig:search-vary-N}
\end{figure*}

%Figure~\ref{fig:search-vary-m} shows the search time for growing pattern lenghts. It shows the case where the pattern exists and where it does not (an extra error is added). The fact that the search is done using an early-exit strategy explains why non existing patterns take longer than existing patterns. The value presented is the number of character comparisons the algorithm makes.

Figure~\ref{fig:search-vary-N} shows searching for a 15~character long sequences on a 2-error dotted tree. Presented are the number of character comparisons performed (averaged over 10\,000 such searches) while varying the text size. We see that after an initial small growth which is explained by the increasing density of the tree, the search time is roughly constant.

### 2005-09-19 paper/search.tex
A submission ready version of the paper

To search a string with at maximum k errors in a k-error dotted tree, we need to descend the tree. Whenever we find a node we follow both the dot link and the edge which matches the current position of the string we are searching. If we have a mismatch in the middle of an edge, we continue, decreasing the amount of errors permitted.

Before introducing the pseudo-code, it is important to understand that suffix trees are normally implemented so that each node consists of only two indices into the original string which are the start and end position of its incoming edge. Figure~\ref{fig:mississippi-nodes} shows an example of this structure (for the now familiar \putstring{mississippi}).

\begin{figure}
\includegraphics[width=\textwidth]{figures/mississippi-nodes}
\caption{Suffix Tree for the string \putstring{mississippi}}%
\label{fig:mississippi-nodes}
\end{figure}

When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$ (as happens in the C~Programming Language).

@@

The search algorithm is presented in Figure~\ref{algo:search}. Reporting of the leafs below a certain node can be done by a Depth First Search reporting any leafs found. If one is interested in knowing only whether the string occurs, then the lines which report the matches can be substituted by early exits returning true.

This algorithm runs in time $\mathcal{O}(m^{k+1})$:

The recursive call at line~\ref{algoline:search:followDot} means that a dot link was followed. There are at most $\mathcal{O}(m^k)$ places where a dot link can be followed while descending a string~\footnote{For $k=1$ each point in the string. For $k>1$, each path for $k-1$, can now have at most $m$ extra branches.}. Following one path in the tree takes $\mathcal{O}(m)$ time. Therefore, the total time is $\mathcal{O}(m^{k+1})$ without the reporting of occurrences (which can be done in time linear to the number of occurrences).

>>>>

To search a string with at maximum k errors in a k-error dotted tree, we descend the tree. Whenever we find a node we follow both the dot link and the edge which matches the current position of the string we are searching. If we find a mismatch in the middle of an edge, we continue, decreasing the amount of errors permitted.
When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$ (as happens in the C~Programming Language).

@@

The process just described is Algorithm~\ref{algo:search}. Reporting of the leafs below a certain node can be done by a Depth First Search. If one is interested in knowing only whether the string occurs, then the lines which report the matches can be substituted by early exits returning true.

This algorithm runs in time $\mathcal{O}(m^{k+1})$: The recursive call at line~\ref{algoline:search:followDot} corresponds to following a dot link. There are at most ${m \choose k}=\mathcal{O}(m^k)$ different paths which must be followed. Following one path in the tree takes $\mathcal{O}(m)$ time. Therefore, the total time is $\mathcal{O}(m^{k+1})$ without the reporting of occurrences (which can be done in time linear to the number of occurrences).

### 2005-09-19 paper/space.tex
A submission ready version of the paper

How much space does a k-error dotted tree take? Suffix Trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). One-error dotted trees have more nodes. The worst case presented in Figure~\ref{fig:aaaa} is a string of the form $a^n = aaa\dots$ which generates a one-dotted tree taking $\mathcal{O}(n^2)$ nodes. For a larger number of errors $k$, it is easy to see that this leads to $\mathcal{O}(n^{k+1})$ nodes. We can do better, however, by considering the expected case.

@@

We can now show that $N_k = \mathcal{O}(nl^k)$ by induction. It is a known fact that $N_0 = \mathcal{O}(n)$. The algorithm for turning a k-error into a (k+1)-error dotted tree, can be looked at the following way~\footnote{We are here considering that the second \textit{foreach} loop of Figure~\ref{algo:construct-k} runs in a specific order which is in fact difficult to code for. However, as an analysis tool, it is a valid assumption.}: First it makes an almost exact copy of the k-error dotted tree and as the error tree at the root and it clears all the other error trees. Then it proceeds in stages, making a (possibly incomplete) copy of this tree spread amongst the nodes at suffix-depth 1. It proceeds to the other nodes in increasing suffix-depths up to $l$. At each such step, the number of nodes is increased by a maximum of $N_k$. Therefore, we start with $N_k$ nodes, make an almost full copy, and copy that at most $l$ times, $N_{k+1}=\mathcal{O}(N_k(l+1))$. Assuming $N_k=\mathcal{O}(nl^k)$ by induction we conclude $N_{k+1}=\mathcal{O}(nl^{k+1})$.

>>>>

Suffix trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). Dotted trees have more nodes. The worst case is a string of the form $a^n = aaa\dots$ which generates a one-dotted tree taking $\mathcal{O}(n^2)$ nodes (illustrated in Figure~\ref{fig:aaaa}). For a larger number of errors $k$, this string leads to $\mathcal{O}(n^{k+1})$ nodes. We can do better, however, by considering the expected case.

@@

We now show $N_k = \mathcal{O}(nl^k)$ using induction. It is a known fact that $N_0 = \mathcal{O}(n)$. The algorithm for turning a k-error into a (k+1)-error dotted tree, can be looked at the following way~\footnote{We are here considering that the second \textit{foreach} loop of Algorithm~\ref{algo:construct-k} runs in a specific order which is in fact difficult to code for. However, as an analysis tool, it is a valid assumption.}: First it makes an almost exact copy of the k-error dotted tree and as the error tree at the root and it clears all the other error trees. Then it proceeds in stages, making a (possibly incomplete) copy of this tree spread amongst the nodes at suffix-depth 1. It proceeds to the other nodes in increasing suffix-depths up to $l$. At each such step, the number of nodes is increased by a maximum of $N_k$. Therefore, we start with $N_k$ nodes, make an almost full copy, and copy that at most $l$ times, $N_{k+1}=\mathcal{O}(N_k(l+1))$. Assuming $N_k=\mathcal{O}(nl^k)$ by induction we conclude $N_{k+1}=\mathcal{O}(nl^{k+1})$.

### 2005-09-19 paper/structure.tex
A submission ready version of the paper

\begin{figure}
\centering
\includegraphics[width=.75\textwidth]{figures/mississippi-0}
\caption{Suffix Tree for the string \putstring{mississippi}}%
\label{fig:mississippi-0}
\end{figure}

@@

Each node in a Patricia tree has a path leading to it which forms a string. If the node $n$ has the leading path $w$, we shall also refer to $n$ as $\overline{w}$one .
\end{definition}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/mississippi-1.eps}
\caption{1-error dotted tree for \putstring{mississippi}}%
\label{fig:mississippi-1}
\end{figure}

@@

>>>>

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{figures/mississippi-0}
\caption{Suffix tree representation for the string \putstring{mississippi}}%
\label{fig:mississippi-0}
\end{figure}

@@

Each node in a Patricia tree has a path leading to it which forms a string. If the node $n$ has the leading path $w$, we shall also refer to $n$ as $\overline{w}$.
\end{definition}

@@

\begin{figure*}
\includegraphics[width=\textwidth]{figures/mississippi-nodes}
\caption{Suffix Tree for the string \putstring{mississippi}}%
\label{fig:mississippi-nodes}
\end{figure*}

It is important to understand that suffix trees are normally implemented so that each node consists of only two indices into the original string which are the start and end position of its incoming edge. Figure~\ref{fig:mississippi-nodes} illustrates this.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/mississippi-1.eps}
\caption{1-error dotted tree for \putstring{mississippi}}%
\label{fig:mississippi-1}
\end{figure*}
